---
title: "Empirical"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---
title: "Empirical vignette V2"
author: "Sigert Ariens"
date: "2025-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The irf_empirical function 

The \code{\link[impulseR]{irf_empirical}} first fits a specific dynamic regression model to the data using the lm() function in R. Then, the observed covariate values, the estimated innovations, and the parameter estimates are passed to the \code{\link[impulseR]{irf_generator}} function to calculate the responses needed for empirical trajectory plots. Some technical details about initial innovations (manuscript appendix X) are handled automatically. To illustrate the use of the \code{\link[impulseR]{irf_empirical}} function, we first use the \code{\link[impulseR]{irf_generator}} to simulate some toy data. 

### Simulate some toy data

We simulate toy data from the following lag-2 ADL:

$$ y_{t} = 1+ 0.5 y_{t-1} - 0.2 y_{t-2} - 2 x_{t} + 0.5 x_{t-1} + x_{t-2}+ v_{t}$$

where

$$v_{t} \overset{iid}{\sim}N(0,0.6)$$

and 

$$x_{t} \overset{iid}{\sim}N(1,1)$$

We simulate the data as conventional simulation studies would simulate data, i.e. by first allowing for a burn in period. The default "burnin = TRUE" need therefore not be changed, but we display the argument in the code below for clairty. 


```{r Simulate some toy data}

xdata <- rnorm(50,-1,1) # Generate some covariate values
vdata <- rnorm(50,0,0.6) # Generate some innovations

toy_data <- irf_generator(1, c(0.5,-0.2), c(-2,0.5,1), x = xdata, innovations = vdata, burnin = TRUE) # set intercept, AR parameters, covariate parameters, x, and v. 

```

The column "irf" contains the model implied values for the dependent variable, $y_{t}$. The covariate values are provided in the column "x". These are the only two columns we have available by measuring the criterion variable and the covariate in practice. For illustrative purposes, we create a new dataframe called "observed data" containing the time index and these two columns. We rename the dependet variable to "Y" and the independent variable to "X":


```{r select the data we would have available}

observed_data <- empirical_data[,c(2,6)]

colnames(observed_data) <- c("Y", "X")

```


### Using the irf_empirical function

#### Arguments

The \code{\link[impulseR]{irf_empirical}} function has two main sets of arguments. The first have to do with the data frame and the columns containing the time series of interest. The second have to do with the order of the ADL model you would like to fit to your data. 

##### Data

To fit an ADL model to data and calculate the system responses, one first provides a dataframe containing the variables of interest. The arguments "data" and "cols" allow one to specify which columns in a dataset should be subject to analysis. The dependent variable should be placed first, the covariate should come second.  

##### Model specification

The ADL(p,q) notation is used to specify the model. The argument "y_lags" is used to specify how many AR effects are allowed for. The argument "x_lags" is used to specify how many lagged covariate effects are allowed for. By default, a contemporaneous covariate effect $\beta_{x}$ will be estimated. To estimate dead start dynamics, the optional argument "contemporaneous" should be set to FALSE. By setting "x_lags = 0", one therefore fits an LCV model to the data. An intercept parameter will always be estimated.

#### Example

To fit the data generating model back to the data, we supply the \code{\link[impulseR]{irf_empirical}} function with the following arguments:

```{r}

estimated_responses <- irf_empirical(data = observed_data, cols = c("Y","X"), y_lags = 2, x_lags = 2 )
```

The estimated_responses object is a list containing 5 elements. 

- A fit object named *fit*. This is the raw output of the lm() fit object. Note that estimates of the first $p$ innovations will not be returned by this fit object, because the initial values $y_{0}, y_{1}$ and $x_{0},x_{1}$ were conditioned on for estimation. 
- A dataframe named *irf*, containing the estimated cumulative responses towards the input variables over the observation period. This output has the same format as the output of the \code{\link[impulseR]{irf_generator}} function. The column estimated_responses\$irf\$innovations contains the first $p$ innovations needed to construct the empirical trajectory plots.;
- A list containing the estimate of the intercept parameter
- A list containing the estimates of the covariate parameters. The first element will be $\hat{\beta}_{x}$, the second element will be $\hat{\beta}_{Lx}$, and so on. The list will be of dimension $q+1$ if $q$ lagged covariate parameters are allowed for and a contemporaneous effect is estimated. 
- A list containing the estimates of the AR parameters, again in order of increasing lag.  

To construct an empirical trajectory plot, we simply call the \code{\link[impulseR]{irf_plot}} function:

```{r}

irf_plot(estimated_responses$irf)

```


#### Estimated parameter values, hypothetical input vectors

The output of the \code{\link[impulseR]{irf_empirical}} function contains the cumulative responses over the study period. Given that we have parameter estimates available, it is straightforward to use the \code{\link[impulseR]{irf_generator}} function, as described in vignette X, to construct meaningful scaled impulse responses. For example, we can calculate the standard deviation of the estimated innovations, and of the covaraite, and use these construct scaled impulse response plots. 

#### Missing values

The \code{\link[impulseR]{irf_empirical}} function handles missing values through listwise deletion. If other methods of estimation are used, the parameter estimates and estimated innovations can be supplied to the \code{\link[impulseR]{irf_generator}} function to produce empirical trajectory plots. If the estimator conditions on initial values, the user would have to fix the first unobserved innovations manually (see Appendix 2 in the manuscript).


#### The burn in option

First, the \code{\link[impulseR]{irf_empirical}} fits the model to data, and only afterwards does it calculate the system responses. When proceeding from the estimation to the calculation step, the \code{\link[impulseR]{irf_empirical}} function sets the burnin option to FALSE. Otherwise, there will be slight discrepancies between $y_{t}^{obs}$ (the observed values of the dependent variable in the data) and $y_{t}$ (the model implied total response). This section illustrates why the burnin option is set to false by default when empirical trajectory plots are desired.

We call the \code{\link[impulseR]{irf_empirical}} function to our empirical data twice, once with the default "burnin = FALSE" option, and one with the default overridden. 

```{r whynotburn}
# Fit the data and do not burn in the intercept response (default behavior)

est_raw <- irf_empirical(empirical_data, cols = c("Y", "X"), y_lags = 2, x_lags = 2)

# Fit the data and burn in the intercept response (not recommended for empirical trajectory plots)
est_burned <- irf_empirical(empirical_data, cols = c("Y", "X"), y_lags = 2, x_lags = 2, burnin = TRUE)

head(observed_data,n=10)
head(est_raw$irf,n=10)
head(est_burned$irf,n=10)


```

The column "Y" in the dataframe "observed data" contains the observed values for the criterion variable, $y_{t}^{obs}$. The column irf" in est_raw contains the model implied values $y_{t}$ if the intercept response is not burned in. The column "irf" in est_burned contains the model implied values $y_{t}$ if the intercept response is burned in.  Even though in reality the data were generated with the intercept response burned in, there are slight discrepancies between the model implied values $y_{t}$ and the values observed in the data, $y_{t}^{obs}$, when the burn in option is set to TRUE, though only for the first few time points. In comparison, the default "burnin = FALSE" option provides a truthful representation of the observed data into estimated latent system responses.
