---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installing and using the package

To install the package, you can use the `remotes` package:

```{r, eval = FALSE}
remotes::install_gitlab("u0133721/impulseR", host = "gitlab.kuleuven.be/")
```

To load the package, use `library` 

```{r}
library(impulseR)
```


## Overview of the package

The `impulseR` package is designed to provide researchers with a user friendly way of visualizing and understanding the behavior implied by a dynamic regression model, by serving as a tool for the calculation and visualization of the system responses implied by the model. The `impulseR` package can be used to study the behavior of any dynamic regression model of the form:

$$\begin{equation}
y_{t} = \alpha + \sum_{i=1}^{p}\phi_{i}y_{t-i} + \beta_{x}x_{t} + \sum_{j=1}^{q}\beta_{L^{j}x}x_{t-j} + v_{t} ,
\end{equation}$$

where $\alpha$ is an intercept parameter, $\phi_{i}$ is the $i$th autoregressive effect, and $\beta_{L^j x}$ is the $j$th lagged covariate effect. 
$p$ and $q$ are the number of AR parameters and lagged covariate parameters respectively. 

The package allows one to calculate different types of _system responses_ (impulse responses, scaled impulse responses, cumulative responses, and total responses). In each case, parameter values (hypothetical or estimated) are supplied to the package functions. Depending on which input vectors one supplies, one of these system responses will be returned. The different types of system responses all rely on the derivation of the _impulse response function_ of the model. The core function -- namely `irf_generator` -- derives the impulse response functions for a dynamic regression model using the analytic techniques discussed in Appendix 1 of Ariens, Vanhasbroeck, Lafit, & Ceulemans (submitted). The function will return different types of system responses depending on the input vectors supplied, as explained in detail in the vignette on [Analysis](https://impulser-5cae6f.pages.gitlab.kuleuven.be/articles/irf_generator.html).

Additionally, the package allows for the construction of empirical trajectory plots (see manuscript, Figure 6). By calculating cumulative responses towards observed covariate values and estimated innovations, we can decompose the empirical data into latent system responses. The `irf_empirical` function performs estimation and processing steps necessary to construct empirical trajectory plots, as detailed in the vignette on [Empirical data](https://impulser-5cae6f.pages.gitlab.kuleuven.be/articles/irf_empirical.html). A plotting function `irf_plot` can furthermore be used for visualization, as detailed in the vignette on [Plotting](https://impulser-5cae6f.pages.gitlab.kuleuven.be/articles/plotting.html).

Currently the package only supports models with a single time varying covariate. 