---
title: "Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(impulseR)
```

In this vignette, we discuss how the `irf_generator` function can be used to generate the different types of system responses discussed in the manuscript. 

## Arguments

`irf_generator` has two main sets of arguments. The first set are the model parameters. The second set are the input vectors we supply the system with. Different system responses will be returned depending on which input vectors are supplied. The `burnin` argument is discussed afterwards.  

### Parameters

The different types of system responses can be calculated by supplying parameter values and input vectors to the `irf_generator` function. We distinguish the autoregressive (AR) parameters $\{ \phi_{1}, \phi_{2},...\}$ from covariate parameters, $\{\beta_{x}, \beta_{Lx},\beta_{L^{2}x},...\}$, and the intercept parameter $\alpha$. Note that the values of the parameters can be hypothetical or estimated. They are supplied using the following arguments:

- `intercept`: A single numeric value that denotes the value of the intercept of the model;
- `x_params`: A numeric vector of length length $1 + q$ containing the values of the covariate parameters $\{ \beta_{x}, \beta_{Lx},\beta_{L^{2}x}, \dots,\beta_{L^{q}x} \}$. Importantly, $q$ is the number of free lagged covariate effects. The first element of the vector is taken to be the contemporaneous effect of the covariate, $\beta_{x}$;
- `ar_params`: A numeric vector of length $p$ containing the values of the AR parameters, $\{ \phi_{1}, \phi_{2}, \dots,\phi_{p} \}$.

Although one can supply arbitrary parameter values, certain restrictions on the values of the AR parameters are required for stability. Specifically, one may expect to see unstable (i.e., exploding) system responses when some of the roots of the lag polynomial $\phi(L): 1 - \sum_{i=1}^{p}\phi_{i}L^{i}$ are inside the complex unit circle. A warning will be returned if this is the case. We advise against using this package for the study of systems which are not stable, since stability is assumed in the derivations. 

### Input vectors

The values of the covariate, $\{ x_{0}, x_{1},...,x_{t} \}$, and the values of the innovations, $\{ v_{0}, v_{1},...,v_{t} \}$ are supplied as vector arguments. The input vector for the intercept is a unit vector which is supplied automatically. Different system responses will be returned depending on which input vectors are supplied. To make this concrete, let us consider the following LCV model with two AR parameters:

$$y_{t} = 2 + 0.7 y_{t-1} -0.5 y_{t-2} + 2 x_{t} + v_{t}$$


#### Impulse response 

Say we want to perturb the covariate by a unit, and trace the effect of this perturbation over time. An impulse response will be returned if a vector of the form $\{1,0,0,...,0\}$ is supplied as input vector. To only retrieve this impulse response, we set the intercept to 0 (or, equally, leave the intercept argument unspecified), and all other input vectors to zero vectors. We call the `irf_generator` function as follows:

```{r}
data <- irf_generator(
  x_params = 2,
  ar_params = c(0.7, -0.5),
  x = c(1, 0, 0, 0, 0, 0, 0, 0, 0)
)
```

The output is a list containing the following entries:

- `intercept`, `ar_params`, `x_params`: The values for the parameters of the model;
- `irf`: A data.frame containing the computed system responses. 

The data.frame contains the results we are interested in, and itself contains the following columns:

- `time`: The time index, starting at $t = 0$;
- `irf`: The total response;
- `irf_intercept`: The cumulative response towards the unit vector; 
- `irf_x`: The response towards the covariate, which in this case is of the form $\{1,0,0,0,...,0 \}$ and therefore equal to the impulse response function ($h_{x}(s)$); 
- `irf_v`: The response towards the innovations, which in this case is $0$;
- `x`: The covariate input vector; 
- `innovations`: The innovation input vector. 

```{r}
data$irf
```

If we had supplied any other vector \code{x} than $\{1,0,0,0,...,0 \}$, the column `irf_x` would contain the appropriate system response.

To visualize the impulse response $h_{x}(s)$, we can provide the `irf_plot` function with the data.frame of system responses:

```{r}
#| fig.alt: >
#|   One sees a plot with three lines, namely a black dotted line that is overlayed on a red solid line, and a green horizontal line.
#|   These lines represent the system responses to a single impulse at time $t = 0$, represented by a red lollipop, where the black line represents the system's total response, the red line represents the system's response to the impulse from the covariate, and the green line represents the system's response to the impulse from the innovations.
#|   While the green line remains $0$ across time, the red line starts out high and decreases rapidly, overshooting the value $0$ only to increase again, showing a sawtooth pattern that converges to $0$.
#| fig.cap: > 
#|   _Visualization of an impulse response function for an LCV model with $\beta_i = 2$, and $\phi_i = \{0.7, -0.5\}$. 
#|   A single unit impulse is given by the covariate at time $t = 0$._
irf_plot(data$irf)
```

We observe the impulse response towards the covariate, $h_{x}(s)$ in red. Note that the plotting function will automatically denote the time index by $s$ and use appropriate legends if the input vectors are of the form $\{1,0,0,0,...,0 \}$ (impulse responses). The plotting function will also automatically plot the model-implied total response $y_{t}$ as a black dotted line. In this case, since we did not present any innovations, the total response is simply `irf_x`, so the black dotted line and the red line overlap (this may be a bit hard to see with the default colors). 

The plotting function allows one to select which responses to visualize, and customize other features (see the vignette on [Plotting](https://impulser-5cae6f.pages.gitlab.kuleuven.be/articles/plotting.html)).


#### Scaled impulse response 

A scaled impulse response will be returned if a vector of the form $\{a,0,0,...,0\}$ is supplied as input vector, where $a$ is a scalar. A useful scalar can be a scalar indicating a standard, or typical deviation in the input variable. For example, if we know that $v_{t} \overset{iid}{\sim} N(0, 2^2)$, then it might be insightful to assess the system response towards a standard deviation increase in the innovations. We supply the function with this information by supplying the appropriate input vector:

```{r}
data <- irf_generator(
  x_params = 2,
  ar_params = c(0.7, -0.5),
  innovations = c(2, 0, 0, 0, 0, 0, 0, 0, 0, 0)
)
```

The output is again a data frame with the same columns as before. The scaled impulse response $h_{v}(s)\sigma_{v}$ is in the column `irf_v`, and now the column `innovations` contains a $2$ at $t = 0$:

```{r}
data$irf
```

Applying the `irf_plot` function yields a visualization of the scaled impulse response:

```{r}
#| fig.alt: >
#|   One again sees a plot with three lines, namely a black dotted line that is overlayed on a green solid line, and a red horizontal line.
#|   These lines again represent the system responses to a single impulse at time $t = 0$, represented by a green lollipop, where the black line represents the system's total response, the red line represents the system's response to the impulse from the covariate, and the green line represents the system's response to the impulse from the innovations.
#|   While the red line remains $0$ across time, the green line starts out high and decreases rapidly, overshooting the value $0$ only to increase again, showing a sawtooth pattern that converges to $0$.
#| fig.cap: > 
#|   _Visualization of a scaled impulse response function for an LCV model with $\beta_i = 2$, and $\phi_i = \{0.7, -0.5\}$. 
#|   A single scaled impulse of size $2$ is given by the innovations at time $t = 0$._
irf_plot(data$irf)
```

We see that now an innovation of magnitude $2$ was supplied (green lollipop). The scaled impulse response $h_{v}(s)\sigma_{v}$ is drawn as a green line. The legend denotes this response as $h_{v}(s) v_0$, where $v_0$ signifies that the curve no longer represents the impulse response, but the response when supplied by a non-unit input variable. Throughout this documentation, red will typically be reserved for covariate related objects and green for innovation related objects (though this can be customized by the user).

#### Cumulative responses

A cumulative response will be returned if the input vector contains non-zero values for arguments beyond the first argument. Let us supply two consecutive unit innovations, followed by a number of $0$ values to show how the system relaxes:

```{r}
data <- irf_generator(
  x_params = 2,
  ar_params = c(0.7, -0.5),
  innovations = c(1, 1, 0, 0, 0, 0, 0, 0, 0, 0)
)
```

The output has the same structure. The cumulative response towards the innovations $(h_{v} * v)_t$ is again given under the column `irf_v`. We see that the column `irf_v` will always contain the system response towards the innovations. 

```{r}
data$irf
```

Applying the `irf_plot` function yields a trajectory plot:

```{r}
#| fig.alt: >
#|   This plot resembles the previous one, albeit with two green lollipops instead of one, representing more than one impulse that has been given to the system (specifically at times $t = 0$ and $t = 1$).
#|   The overall pattern remains the same, although the green line is pushed further up at $t = 1$ due to the second impulse, only to relax after this timepoint.
#| fig.cap: > 
#|   _Visualization of a cumulative response function for an LCV model with $\beta_i = 2$, and $\phi_i = \{0.7, -0.5\}$. 
#|   Two unit impulses are given by the innovations at times $t = 0$ and $t = 1$._
irf_plot(data$irf)
```

We see that the time index has changed to $t$, and the legend denotes the cumulative response towards the innovations by $(h_{v} * v)_t$.

#### Total response

As an example of a total response, we will supply a unit covariate input and a unit innovation at the same time. We supply the `irf_generator` function with both input vectors and visualize the trajectory:

```{r}
#| fig.alt: >
#|   This plot resembles the previous ones, but now a single unit impulse has been provided by both the covariate and the innovations at time $t = 1$.
#|   Besides the original three lines, a fourth, gray horizontal line has appeared at $0$, representing the effect of the intercept.
#|   The overall pattern remains the same as before; All three lines show a sawtooth pattern of convergence to $0$.
#| fig.cap: > 
#|   _Visualization of the total response as the sum of the covariate and innovation responses functions for an LCV model with $\beta_i = 2$, and $\phi_i = \{0.7, -0.5\}$. 
#|   Two unit impulses are given at time $t = 0$; One through the covariate and one through the innovations._
data <- irf_generator(
  x_params = 2,
  ar_params = c(0.7, -0.5),
  x = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  innovations = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0)
)

irf_plot(data$irf)
```


## Intercept parameters

To include an intercept parameter in the visualization of system responses, one can supply the intercept parameter value to the `irf_generator` function:

```{r}
#| fig.alt: >
#|   This plot resembles the previous one, but now the gray horizontal line (intercept response) and the black dashed line (total response) have been moved up vertically with $2$ units. 
#| fig.cap: > 
#|   _Visualization of the effect of the intercept on the total response.
#|   An LCV model was used with $\alpha = 2$, $\beta_i = 2$, and $\phi_i = \{0.7, -0.5\}$. 
#|   As in the previous figure, two unit impulses are given at time $t = 0$; One through the covariate and one through the innovations._
data <- irf_generator(
  intercept = 2,
  x_params = 2,
  ar_params = c(0.7, -0.5),
  x = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  innovations = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0)
)

irf_plot(data$irf)
```

By default, the `irf_generator` will 'burn in' the part of the system response due to the intercept parameter analytically, in effect fixing it to its model-implied time limit. This allows one to interpret the other system responses relative to the hypothetical equilibrium state of the system. 

However, one can override this default behavior by setting the argument `burnin = FALSE`:

```{r}
#| fig.alt: >
#|   This plot resembles the previous one, but now the gray line (intercept response) also shows sawtooth-like behavior. 
#| fig.cap: > 
#|   _Visualization of the effect of the intercept on the total response when `burnin = FALSE`, that is when the system is not in equilibrium.
#|   As one can see, the impulses provided through the intercept show a sawtooth pattern, showing behavior that is dependent on the initial value._
data <- irf_generator(
  intercept = 2,
  x_params = 2,
  ar_params = c(0.7, -0.5),
  x = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  innovations = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  burnin = FALSE
)

irf_plot(data$irf)
```

By setting the argument `burnin = FALSE`, the behavior of the system is dependent on its initial value while the system approaches its equilibrium state. 

## General dynamic regression models

In the manuscript we focused on lag-1 dynamic regression models for accessibility. Here, we provide some examples of how dynamic regression models with more general lag structures can be investigated using the `irf_generator` function. The following table shows how different parameter settings can be supplied to the `irf_generator` function, together with the associated model: 

 **Abbreviation**   | **`intercept`** | **`ar_params`** | **`x_params`** | **Model**
-------------|----------------------------|--------------------------|----------------------|------
 LR          | `0`                        | `0`                      | `1`                  | $y_{t} = x_{t} + v_{t}$            
 AR(1)       | `1`                        | `0.7`                    | `0`                  | $y_{t} = 1 + 0.7 y_{t-1} + v_{t}$            
 LCV(1)      | `0`                        | `0.7`                    | `1`                  | $y_{t} = 0.7 y_{t-1} + x_{t} + v_{t}$ 
 CF(2)        | `0`                       | `c(0.5, -0.5)`           | `c(2, -1, 1)`        | $y_{t} = 0.5 y_{t-1} - 0.5 y_{t-2} + 2 x_{t} - x_{t-1} + x_{t-2}+ v_{t}$
 ADL(2, 2)   | `3`                        | `c(0.7, 0.2)`            | `c(2, 0.5, 1)`       | $y_{t} = 3+ 0.7 y_{t-1} + 0.2 y_{t-2} + 2 x_{t} + 0.5 x_{t-1} + x_{t-2}+ v_{t}$


Parentheses indicate the number of AR parameters $p$ and lagged covariate parameters $q$. For example, the ADL(2,2) allows for $p = 2$ AR parameters ($\phi_{1}$ and $\phi_{2}$) and for $q = 2$ lagged covariate parameters ($\beta_{Lx}$ and $\beta_{L^{2}x}$). Note that the CF model only has a single value in parenthesis, because $p$ must equal $q$ for the CF restrictions to hold. The CF restrictions are satisfied if $\beta_{L^{j}x} = - \beta_{x}\phi_{j}$ for all $j in \{ 1,...,q \}$ (see Appendix A2 in Ariens et al., submitted). Similarly, the LCV model only has a single value in parenthesis, because the order of a LCV model is determined by the number of free AR parameters, that is by $p$. Indeed, for the LCV model all lagged covariate parameters are fixed to 0. We could refer to a LCV model as an ADL(p, 0) model. Finally, the LR model never requires parentheses, because all lagged parameters are fixed to 0. 

To illustrate how the `irf_generator` function can be used to investigate models with more general lag structures, let us visualize the impulse responses of the CF(2) model in the above table. We know from the manuscript that for any CF model, the effect of the covariate will be purely contemporaneous. We can verify that this is the case for our CF(2) model by calculating the impulse response functions:

```{r}
data <- irf_generator(
  intercept = 0,
  x_params = c(2, -1, 1),
  ar_params = c(0.5, -0.5),
  x = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  innovations = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0)
)
```

We call the `irf_plot` function:

```{r}
#| fig.alt: >
#|   This plot shows the familiar four lines, as well as a lollipop representing the covariate and innovation impulses.
#|   After the covariate impulse (covariate response), the red line initially starts out high but then immediately falls to and remains at $0$ for the remainder of the time.
#|   The green line (innovation response), on the other hand, shows a sawtooth-like pattern, which is nicely followed and overlayed by the black dashed line (total response).
#| fig.cap: > 
#|   _Visualization of the impulse response function of a CV model with $\alpha = 0$, $\beta_i = \{2, -1, 1 \}$, and $\phi_i = \{0.5, -0.5 \}$.
#|   As one can see, the covariate impulse dies out immediately while the innovation impulse persists across time._
irf_plot(data$irf)
```

Indeed, the impulse response $h_{x}(s)$ decays immediately following a perturbation. 
